### 2015-07-07 2015-compmorphon-morphver/poster.tex
some work on the loanword clusters
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@61189 72bbbca6-d526-0410-a7d9-f06f51895060

		Descriptions of a language's
		\begin{itemize}
			\item \textbf{morphotactics} — the morphemes of a language and how they can be combined
			\item \textbf{morphophonology} — the alternations between the various phonological and orthographical forms of each morpheme in a language
		\end{itemize}

@@

		\htwo{Framework:\ HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
		\htwo{Approach}
		\begin{itemize}
			\item morphotactics implemented in \texttt{lexc}
			\item morphophonology implemented in \texttt{twol}
			\item compiled separately; compose-intersected to single transducer\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\hilitegrey{>}\archiphon{D}\archiphon{I}\hilitegrey{>}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\hilitegrey{>}\archiphon{I}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		\end{itemize}

@@

		\headerbox{Example: Tuvan vowel harmony after consonant clusters in loanwords}{name=otherexamples,column=1,span=2,below=tuvanvelars}{
			\begin{minipage}{0.5\textwidth}
				\codeex{"\{I\} harmony"\\
\%\{I\%\}:Vy <=> :Vx [ :Cns* :RealCns ]/[ :0 | \%- ]* \_ ;\\
\-~~~~~~~~~where Vx in ( ү и е э ө а о ы у я ё ю )\\
\-~~~~~~~~~~~~~~ Vy in ( ү и и и ү ы у ы у ы у у )\\
\-~~~~~~~~~matched ;}
			\end{minipage}

			\begin{minipage}{0.5\textwidth}
				\codeex{"\{I\} harmony"\\
\%\{I\%\}:Vy <=> :Vx [ :Cns* :RealCns ]/[ :0 | \%- ]* \_ ;\\
\-~~~~~~~~~except\\
\-~~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь: :Cns* :RealCns ]/:0* \_ ;\\
\-~~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь:0 ]/:[ :0 - ь: ]* \_ ;\\
\-~~~~~~~~~where Vx in ( ү и е э ө а о ы у я ё ю )\\
\-~~~~~~~~~~~~~~ Vy in ( ү и и и ү ы у ы у ы у у )\\
\-~~~~~~~~~matched ;\\
\\
"\{I\} always front when intervening Cль"\\
\%\{I\%\}:и <=> [ :BackVow :Cns* :Cns л: ь: :Cns* :RealCns ]/:0* \_ ;\\
\-~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь:0 ]/:[ :0 - ь: ]* \_ ;}
			\end{minipage}

>>>>

		\htwo{What they describe} %Describe a language's
		\begin{itemize}
			\item \textbf{morphotactics} — the morphemes of a language and how they can be combined
			\item \textbf{morphophonology} — the alternations between the various phonological and orthographical forms of each morpheme% in a language
		\end{itemize}

@@

		\htwo{Framework:\ HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
		\htwo{Approach}
		\begin{itemize}
			\item morphotactics implemented in \texttt{lexc}
			\item morphophonology implemented in \texttt{twol} (SPE-style rules)
			\item compiled separately; compose-intersected to single transducer\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\hilitegrey{>}\archiphon{D}\archiphon{I}\hilitegrey{>}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\hilitegrey{>}\archiphon{I}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		\end{itemize}

@@

		\headerbox{Example: Tuvan vowel harmony after consonant clusters in loanwords}{name=otherexamples,column=1,span=2,below=tuvanvelars}{
			\begin{multicols}{2}
				\htwo{Tuvan vowel harmony}
					\begin{itemize}
						\item High and low vowels harmonise in backness
						\item High vowels also harmonise in roundness
					\end{itemize}
				\htwo{Initial implementation (high-vowels)}

				\begin{minipage}{0.5\textwidth}
					\codeex{"\{I\} harmony"\\
\%\{I\%\}:Vy <=> :Vx [ :Cns* :RealCns ]/[ :0 | \%- ]* \_ ;\\
\-~~~~~~~~~where Vx in ( ү и е э ө а о ы у я ё ю )\\
\-~~~~~~~~~~~~~~ Vy in ( ү и и и ү ы у ы у ы у у )\\
\-~~~~~~~~~matched ;}
				\end{minipage}

				\htwo{Unanalysed forms in corpus}
					\begin{itemize}
						\item forms of loanwords ансамбль `ensemble' and рубль `rouble':\\\
						\-\hfill{}the following vowel is always front unround
					\end{itemize}
					\begin{minipage}[t]{0.5\textwidth}
					\outputex{
						\begin{multicols}{3}
							%790 рубль\\
							347 рубльди\\
							99 рубльге\\
							83 ансамблиниң\vphantom{у}\\
							60 ансамбли\vphantom{у}\\
							54 рубльдиң\\
							28 рубльден\\
							17 ансамбльдиң\vphantom{у}\\
							15 ансамбльдер\vphantom{у}\\
							14 ансамбль\vphantom{у}\\
							11 ансамблин\vphantom{у}\\
							10 ансамблинге\vphantom{у}\\
							8 ансамбльди\vphantom{у}\\
							%2 рубльга\\
							1 рубльдерни\\
							1 рублин\\
							1 рубли\\
						\end{multicols}
					}
					\end{minipage}

					\columnbreak
				\htwo{Revised generalisation}
					\begin{itemize}
						\item After cons.\ cluster ending in ль, vowel always front unround
						\item New discovery?  Appears to be not previously documented
					\end{itemize}

				\htwo{Revised implementation}
				\begin{minipage}{0.5\textwidth}
					\codeex{\hilitegrey{"\{I\} harmony"\\
\%\{I\%\}:Vy <=> :Vx [ :Cns* :RealCns ]/[ :0 | \%- ]* \_ ;}\\
\-~~~~~~~~~except\\
\-~~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь: :Cns* :RealCns ]/:0* \_ ;\\
\-~~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь:0 ]/:[ :0 - ь: ]* \_ ;\\
\-~~~~~~~~~\hilitegrey{where Vx in ( ү и е э ө а о ы у я ё ю )\\
\-~~~~~~~~~~~~~~ Vy in ( ү и и и ү ы у ы у ы у у )\\
\-~~~~~~~~~matched ;}\\
\\
"\{I\} always front when intervening Cль"\\
\%\{I\%\}:и <=> [ :BackVow :Cns* :Cns л: ь: :Cns* :RealCns ]/:0* \_ ;\\
\-~~~~~~~~~~~~[ :BackVow :Cns* :Cns л: ь:0 ]/:[ :0 - ь: ]* \_ ;}
				\end{minipage}
			\end{multicols}
